\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-19 submission/manuscript.tex
Pipeline optimization.

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on macrophyte
epiphytic communities did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities sampled from the surfaces of the seagrass
\emph{Cymodocea nodosa} and invasive macroalga \emph{Caulerpa
cylindracea}. Leaves and thalli were sampled in a meadow of
\emph{Cymodocea nodosa} invaded by the invasive \emph{Caulerpa
cylindracea} and in a monospecific settlement of \emph{Caulerpa
cylindracea} located in the proximity of the meadow at monthly
intervals. For comparison the ambient prokaryotic plankton community was
also characterized. Sequencing results at the OTU level showed a clear
differentiation between ambient water and epiphytic communities and a
host-specific community assemblage. In addition, successional changes
were observed that could be connected to the macrophyte growth cycle.
Taxonomic analysis showed similar high rank groups in the ambient water
and epiphytic communities, with the exception of \emph{Desulfobacterota}
that were found only on \emph{Caulerpa cylindracea}. Only
\emph{Cyanobacteria} showed seasonal change, while other high rank taxa
were present throughout the year. In every analyzed high rank taxa,
phylogenetic groups present throughout the year and comprising most of
the sequences could be identified together with low proportion taxa
showing seasonal patterns connected to the macrophyte growth cycle.
Taken together, epiphytic microbial communities of the seagrass
\emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa cylindracea}
appear to be host-specific and contain taxa that undergo successional
changes.

@@

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were
described (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016). Recently, it was shown that even different morphological niches
within the same alga had a higher influence on bacterial community
variation than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While there is high community variation between host
species is was observed that the majority of metagenome determined
functions were conserved both between host species and individuals
(Burke \emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). This
discrepancy between taxonomic and functional composition could be
explained by the lottery hypothesis. It postulates that an initial
random colonization step is performed from a set of functionally
equivalent taxonomic groups resulting in taxonomically different
epiphytic communities sharing a core set of functional genes (Burke
\emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). In addition,
some of the variation in the observed data could be attributed to
different techniques used in various studies, such as different
protocols for epiphytic cell detachment and/or DNA isolation, as no
standard protocol to study epiphytic communities was established
(Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.}, submitted).

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke \emph{et al.}, 2011b; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal changes
were taken into account low temporal frequency and/or methodologies that
do not allow for high taxonomic resolution were used (Tujula \emph{et
al.}, 2010; Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013;
Michelou \emph{et al.}, 2013). In the present study we describe the
seasonal dynamics of bacterial and archaeal communities on the surfaces
of the seagrass \emph{Cymodocea nodosa} and siphonous macroalgae
\emph{Caulerpa cylindracea} determined on a mostly monthly scale.
Bacterial and archaeal epiphytes were sampled in a meadow of \emph{C.
nodosa} invaded by the invasive \emph{C. cylindracea} and in a locality
of only \emph{C. cylindracea} located in the proximity of the meadow. In
addition, for comparison, the community of the ambient seawater was
characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,408 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,736 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,408
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,114
different OTUs that ranged from 379 to 2,016 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (invaded {[}Number of OTUs, 1,676.9 ±
140.3 OTUs{]} and noninvaded {[}Number of OTUs, 1,745.9 ± 156.3
OTUs{]}), middle vales for \emph{C. nodosa} (Number of OTUs, 1,059.5 ±
212.3 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 530.8 ± 143.1 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (invaded and
noninvaded) communities were characterized by slightly larger values in
Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs made also a high proportion of total sequences (40.3 -- 52.4
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.55, \emph{p} \textless{} 0.001),
while communities from the surfaces of \emph{C. cylindracea} indicated a
non so strongly supported, as in previous cases, separation between
Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31, \emph{p}
\textless{} 0.01) (\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by bacterial (99.1 ± 2.1 \si{\percent}) over
archaeal sequences (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{C. nodosa} were consisted of same groups with the
addition of \emph{Planctomycetota} whose contribution was higher in
summer 2018. In addition, communities from invaded and noninvaded
\emph{C. cylindracea} were similar and characterized by same groups as
seawater and \emph{C. nodosa} communities with the addition of
\emph{Desulfobacterota} (\autoref{community}). Larger differences
between environments and host species could be observed at lower
taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (invaded {[} (7.7 ± 3.9 \si{\percent}){]} and
noninvaded {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.5 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} were comprising, on average,
19.2 ± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly
to \emph{Cyanobacteria}, large differences in the taxonomic composition
between seawater and surface associated communities were found
(\autoref{bactero}). The seawater community was characterized by the NS4
and NS5 marine groups, uncultured \emph{Cryomorphaceae}, uncultured
\emph{Flavobacteriaceae}, NS11-12 marine group, \emph{Balneola},
uncultured \emph{Balneolaceae} and \emph{Formosa}. In contrast, in
surface associated communities \emph{Lewinella}, \emph{Portibacter},
\emph{Rubidimonas}, no relative \emph{Saprospiraceae}, uncultured
\emph{Saprospiraceae}, no relative \emph{Flavobacteriaceae} and
uncultured \emph{Rhodothermaceae} were found. Some groups showed slight
seasonal changes such as no relative \emph{Flavobacteriaceae} that were
more pronounced from November 2017 until June 2018. In contrast,
uncultured \emph{Rhodothermaceae} showed higher proportions from June
2018 until the end of the study period. Surface associated
\emph{Bacteroidota} communities were very diverse as could be observed
in the the high proportion of taxa that grouped as other
\emph{Bacteroidota} (\autoref{bactero}).

@@

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.6 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characterisitic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{C. cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{C. nodosa} communities were consisted of only
0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{C. cylindracea} communities their proportion was 25.7 ±
11.2 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Surfaces of marine macrophytes are harboring biofilms consisted of
diverse microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). No standard protocol has been developed to study these
macophyte associated microbes (Ugarelli \emph{et al.}, 2019). Different
procedures of microbial cells removal from host surfaces were described,
such as host tissue shaking (Nõges \emph{et al.}, 2010), scraping (Uku
\emph{et al.}, 2007) and ultrasonication (Cai \emph{et al.}, 2014). All
these methods showed different removal efficiencies but none was
enabling a complete removal of attached microbial cells. In the present
study, we applied an earlier developed removal protocol (Korlević
\emph{et al.}, submitted), based on a previous idea of direct cellular
lysis (Burke \emph{et al.}, 2009), to ensure an almost complete cell
detachment. The application of a direct lysis procedure coupled with a
high frequency sampling protocol and Illumina high resolution amplicon
sequencing has enabled us to make a detailed description of bacterial
and archaeal communities associated with the surfaces of two marine
macrophytes, \emph{C. nodosa} and \emph{C. cylindracea}.

In the present study, highest richness values were observed for \emph{C.
cylindracea} (invaded and noninvaded), middle for \emph{C. nodosa} and
lowest for seawater derived communities. Higher values for seagrass
associated communities in comparison to seawater were described earlier
and could be attributed to a larger set of inhabitable microniches
existing on macrophyte surfaces (Ugarelli \emph{et al.}, 2019). In
addition, highest values observed for \emph{C. cylindracea} are probably
a consequence of part of epiphytic OTUs that are in contact with the
sediment. \emph{C. cylindracea} stolon is attached to the sediment
surface with rhizoids, so the stolon and rhizoids are in a direct
contact with the sediment. Part of the surface attached \emph{Caulerpa
cylnidracea} community is therefore comprised of OTUs that are in
contact with a different environment and that could cause the observed
increase in richness. In addition, seasonal richness differences
observed for surface attached communities showed slightly higher values
in spring and summer. This pattern could be explained by a higher
macrophyte growth in these seasons (M. Najdek, personal communication;
Zavodnik \emph{et al.}, 1998; Ruitton \emph{et al.}, 2005). During
active periods macrophytes exhibit a more dynamic chemical interaction
with the surface community probably causing an increase in the number of
inhabitable microniches (Borges and Champenois, 2015; Rickert \emph{et
al.}, 2016).

>>>>

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on macrophyte
epiphytic communities did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities sampled from the surfaces of the seagrass
\emph{Cymodocea nodosa} and invasive macroalga \emph{Caulerpa
cylindracea}. Leaves and thalli were sampled in a meadow of
\emph{Cymodocea nodosa} invaded by the invasive \emph{Caulerpa
cylindracea} and in a monospecific settlement of \emph{Caulerpa
cylindracea} located in the proximity of the meadow at monthly
intervals. For comparison the ambient prokaryotic plankton community was
also characterized. Sequencing results at the OTU level showed a clear
differentiation between ambient water and epiphytic communities and a
host-specific community assemblage. In addition, successional changes
were observed that could be connected to the macrophyte growth cycle.
Taxonomic analysis showed similar high rank groups in the ambient water
and epiphytic communities, with the exception of \emph{Desulfobacterota}
that were found only on \emph{Caulerpa cylindracea}. Only
\emph{Cyanobacteria} showed seasonal change, while other high rank taxa
were present throughout the year. In every analyzed high rank taxa,
phylogenetic groups present throughout the year comprised most of the
sequences and could be identified together with low proportion taxa
showing seasonal patterns connected to the macrophyte growth cycle.
Taken together, epiphytic microbial communities of the seagrass
\emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa cylindracea}
appear to be host-specific and contain taxa that undergo successional
changes.

@@

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were
described (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016). Recently, it was shown that even different morphological niches
within the same alga had a higher influence on bacterial community
variation than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While there is high community variation between host
species it was observed that the majority of metagenome determined
functions were conserved both between host species and individuals
(Burke \emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). This
discrepancy between taxonomic and functional composition could be
explained by the lottery hypothesis. It postulates that an initial
random colonization step is performed from a set of functionally
equivalent taxonomic groups resulting in taxonomically different
epiphytic communities sharing a core set of functional genes (Burke
\emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). In addition,
some of the variation in the observed data could be attributed to
different techniques used in various studies, such as different
protocols for epiphytic cell detachment and/or DNA isolation, as no
standard protocol to study epiphytic communities was established
(Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.}, submitted).

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke \emph{et al.}, 2011b; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal changes
were taken into account low temporal frequency and/or methodologies that
do not allow high taxonomic resolution were used (Tujula \emph{et al.},
2010; Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013). In the present study we describe the seasonal
dynamics of bacterial and archaeal communities on the surfaces of the
seagrass \emph{Cymodocea nodosa} and siphonous macroalgae \emph{Caulerpa
cylindracea} determined on a mostly monthly scale. Bacterial and
archaeal epiphytes were sampled in a meadow of \emph{C. nodosa} invaded
by the invasive \emph{C. cylindracea} and in a locality of only \emph{C.
cylindracea} located in the proximity of the meadow. In addition, for
comparison, the community of the ambient seawater was characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,409 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,691 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,409
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,071
different OTUs that ranged from 386 to 2,009 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,679.5 ±
127.4 OTUs{]} and monospecific {[}Number of OTUs, 1,736.3 ± 156.5
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,059 ±
216.2 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 528.2 ± 135.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (39.9 -- 51.8
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.53, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.30, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by bacterial (99.1 ± 2.1 \si{\percent}) over
archaeal sequences (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{C. nodosa} consisted of same groups with the
addition of \emph{Planctomycetota} whose contribution was higher in
summer 2018. In addition, communities from mixed and monospecific
\emph{C. cylindracea} were similar and characterized by same groups as
seawater and \emph{C. nodosa} communities with the addition of
\emph{Desulfobacterota} (\autoref{community}). Larger differences
between environments and host species could be observed at lower
taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (mixed {[} (7.7 ± 3.9 \si{\percent}){]} and
monospecific {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.4 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} were comprising, on average,
19.2 ± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly
to \emph{Cyanobacteria}, large differences in the taxonomic composition
between seawater and surface associated communities were found
(\autoref{bactero}). The seawater community was characterized by the NS4
and NS5 marine groups, uncultured \emph{Cryomorphaceae}, uncultured
\emph{Flavobacteriaceae}, NS11-12 marine group, \emph{Balneola},
uncultured \emph{Balneolaceae} and \emph{Formosa}. In contrast, in
surface associated communities \emph{Lewinella}, \emph{Portibacter},
\emph{Rubidimonas}, no relative \emph{Saprospiraceae}, uncultured
\emph{Saprospiraceae}, no relative \emph{Flavobacteriaceae} and
uncultured \emph{Rhodothermaceae} were found. Some groups showed slight
seasonal changes such as no relative \emph{Flavobacteriaceae} that were
more pronounced from November 2017 until June 2018. In contrast,
uncultured \emph{Rhodothermaceae} showed higher proportions from June
2018 until the end of the study period. Surface associated
\emph{Bacteroidota} communities were very diverse as could be observed
in the high proportion of taxa that grouped as other \emph{Bacteroidota}
(\autoref{bactero}).

@@

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.6 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicoccus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
mixed and monospecific \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characteristic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{C. cylindracea}. On
average they comprised 11.2 ± 13.3 \si{\percent} of all sequences.
Seawater and \emph{C. nodosa} communities consisted of only 0.1 ± 0.08
\si{\percent} and 1.0 ± 0.7 \si{\percent} \emph{Desulfobacterota}
sequences, respectively. In the mixed and monospecific \emph{C.
cylindracea} communities their proportion was 25.7 ± 11.2 \si{\percent}
and 24.0 ± 4.3 \si{\percent}, respectively (\autoref{desulfo}). The
community consisted mainly of no relative \emph{Desulfobacteraceae},
\emph{Desulfatitalea}, no relative \emph{Desulfobulbaceae},
\emph{Desulfobulbus}, no relative \emph{Desulfocapsaceae},
\emph{Desulfopila}, \emph{Desulforhopalus}, \emph{Desulfotalea},
SEEP-SRB4 and uncultured \emph{Desulfocapsaceae} (\autoref{desulfo}).

@@

Surfaces of marine macrophytes harbor biofilms consisting of diverse
microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et al.},
2019). No standard protocol has been developed to study these macophyte
associated microbes (Ugarelli \emph{et al.}, 2019). Different procedures
for removal of microbial cells from host surfaces were described, such
as host tissue shaking (Nõges \emph{et al.}, 2010), scraping (Uku
\emph{et al.}, 2007) and ultrasonication (Cai \emph{et al.}, 2014). All
these methods showed different removal efficiencies but none was
enabling a complete removal of attached microbial cells. In the present
study, we applied an earlier developed removal protocol (Korlević
\emph{et al.}, submitted), based on a previous idea of direct cellular
lysis (Burke \emph{et al.}, 2009), to ensure an almost complete cell
detachment. The application of a direct lysis procedure coupled with a
high frequency sampling protocol and Illumina high resolution amplicon
sequencing has enabled us to make a detailed description of bacterial
and archaeal communities associated with the surfaces of two marine
macrophytes, \emph{C. nodosa} and \emph{C. cylindracea}.

In the present study, highest richness values were observed for \emph{C.
cylindracea} (mixed and monospecific), middle for \emph{C. nodosa} and
lowest for seawater derived communities. Higher values for seagrass
associated communities in comparison to seawater were described earlier
and could be attributed to a larger set of inhabitable microniches
existing on macrophyte surfaces (Ugarelli \emph{et al.}, 2019). In
addition, highest values observed for \emph{C. cylindracea} are partly
due to its contact with the sediment. \emph{C. cylindracea} stolon is
attached to the sediment surface with rhizoids, so the stolon and
rhizoids are in a direct contact with the sediment. In addition,
seasonal richness differences observed for surface attached communities
showed slightly higher values in spring and summer. This pattern could
be explained by a higher macrophyte growth in these seasons (M. Najdek,
personal communication; Zavodnik \emph{et al.}, 1998; Ruitton \emph{et
al.}, 2005). During active periods macrophytes exhibit a more dynamic
chemical interaction with the surface community probably causing an
increase in the number of inhabitable microniches (Borges and
Champenois, 2015; Rickert \emph{et al.}, 2016).
